New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-16 jso.qmd
Typos

The index properties, smoothness and squintability, numerically characterise the complexity of projection pursuit index, and hence potential optimisation problems. @tbl-smoothness-squintability presents the parameters for calculating both metrics estimated from the Gaussian process (variance, range, smooth, and nugget) and the scaled logistic function ($\theta_1$ to $\theta_4$) for each case considered in the simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric.

@@

The simulation study demonstrated the that these metrics can be used to assess a PP index, and compare optimisers. Here we explain how they can be used for new index and optimiser development. Using the JSO optimiser for GTPP requires a change in user behaviour. For all the existing optimisers a single optimisation path is followed. The JSO has multiple optimisation paths, and needs additional tools to incorporate into the PPGT, as explained here.

>>>>

The index properties, smoothness and squintability, numerically characterise the complexity of projection pursuit index, and hence potential optimisation problems. @tbl-smoothness-squintability presents the parameters for calculating both metrics estimated from the Gaussian process (variance, range, smooth, and nugget) and the scaled logistic function ($\theta_1$ to $\theta_4$) for each case considered in the simulation. The column $\nu$ is used as the smoothness metric and the
column $\varsigma$ is calculated as equation (@eq-squintability-parametric)
as the squintability metric.

@@

The simulation study demonstrated that these metrics can be used to assess a PP index, and compare optimisers. Here we explain how they can be used for new index and optimiser development. Using the JSO optimiser for PPGT requires a change in user behaviour. For all the existing optimisers a single optimisation path is followed. The JSO has multiple optimisation paths, and needs additional tools to incorporate into the PPGT, as explained here.

### 2024-09-16 jso.tex
Typos

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, jellyfish search optimiser
(JSO), optimisation, grand tour, high-dimensional data, exploratory data
analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}

\setstretch{2}
\section{Introduction}\label{introduction}

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, breakable, colback=white, toptitle=1mm, titlerule=0mm, left=2mm, arc=.35mm, bottomrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}]

@@

\section{Assessing optimisers: a simulation study}\label{sec-sim-deets}

@@

\end{figure}%

\section{Results}\label{sec-sim-res}

\subsection{Performance of JSO relative to
CRS}\label{performance-of-jso-relative-to-crs}

@@

\subsection{Factors affecting JSO success rate: index properties and
jellyfish
hyper-parameters}\label{factors-affecting-jso-success-rate-index-properties-and-jellyfish-hyper-parameters}

@@

\end{figure}%

The index properties, smoothness and squintability, numerically
characterise the complexity of projection pursuit index, and hence
potential optimisation problems.
Table~\ref{tbl-smoothness-squintability} presents the parameters for
calculating both metrics estimated from the Gaussian process (variance,
range, smooth, and nugget) and the scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column ``smooth'' is used as the smoothness metric and
the column ``squint'' is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

\begin{table}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (outputscale \(\eta\), lengthscale \(\ell\),
smoothness \(\nu\), and nugget \(\sigma\)) and scaled logistic function
(\(\theta_1\) to \(\theta_4\) and \(\varsigma\)) for the pipe-finding
and sine-wave finding problems. The columns \(\nu\) and \(\varsigma\)
represent the smoothness and squintability measures respectively.}

\centering{

@@

\begin{tabular}{|>{}lcc>{}c|ccc>{}c|cccc>{}c|}
\toprule
  & shape & index & d & $\eta$ & $\ell$ & $\nu$ & $\sigma$ & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & $\varsigma$\\
\midrule
1 & pipe & holes & 6 & 0.04 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.01 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.01 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.01 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.13 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.13 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.35 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.35 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor & 6 & 0.18 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess & 6 & 0.28 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines & 6 & 0.20 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

\setlength{\LTpost}{0mm}

\begin{longtable}{lcc}

\caption{\label{tbl-mod-output}Jellyfish success rate relative to index
properties and jellyfish hyper-parameters. This is the summary from a
logistic regression fit to smoothness, squintability, dimension, running
time, number of jellyfish and maximum number of tries. Interestingly,
squintability and dimension strongly affect jellyfish optimisation
success. The number of jellies marginally affects success, but index
smoothness, running longer and increasing the number of tries do not.}

\tabularnewline

\toprule
\textbf{Characteristic} & \textbf{OR} \textbf{(95\% CI)}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Smoothness & 3.30 (0.08 to 274) & 0.533 \\ 
Squintability & 7.85 (2.40 to 37.7) & \textbf{0.003} \\ 
Dimension & 0.53 (0.30 to 0.83) & \textbf{0.014} \\ 
Long runtime & 0.42 (0.03 to 5.44) & 0.497 \\ 
Number of jellyfish & 1.24 (0.98 to 1.63) & \textbf{0.089} \\ 
Maximum number of tries & 1.12 (0.84 to 1.52) & 0.452 \\ 
\bottomrule

\end{longtable}

\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}OR = Odds Ratio, CI = Confidence Interval\\
\end{minipage}

\section{Practicalities}\label{sec-discussion}

The simulation study demonstrated the that these metrics can be used to
assess a PP index, and compare optimisers. Here we explain how they can
be used for new index and optimiser development. Using the JSO optimiser
for GTPP requires a change in user behaviour. For all the existing
optimisers a single optimisation path is followed. The JSO has multiple
optimisation paths, and needs additional tools to incorporate into the
PPGT, as explained here.

\subsection{Using the JSO in a PPGT}\label{using-the-jso-in-a-ppgt}

@@

\section*{Supplementary materials}\label{supplementary-materials}
\addcontentsline{toc}{section}{Supplementary materials}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{anyfontsize}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, jellyfish search optimiser
(JSO), optimisation, grand tour, high-dimensional data, exploratory data
analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

\setstretch{2}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, breakable, coltitle=black, opacitybacktitle=0.6, arc=.35mm, toprule=.15mm, left=2mm, bottomtitle=1mm, leftrule=.75mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}]

@@

\hypertarget{sec-sim-deets}{%
\section{Assessing optimisers: a simulation study}\label{sec-sim-deets}}

@@

\end{figure}

\hypertarget{sec-sim-res}{%
\section{Results}\label{sec-sim-res}}

\hypertarget{performance-of-jso-relative-to-crs}{%
\subsection{Performance of JSO relative to
CRS}\label{performance-of-jso-relative-to-crs}}

@@

\hypertarget{factors-affecting-jso-success-rate-index-properties-and-jellyfish-hyper-parameters}{%
\subsection{Factors affecting JSO success rate: index properties and
jellyfish
hyper-parameters}\label{factors-affecting-jso-success-rate-index-properties-and-jellyfish-hyper-parameters}}

@@

\end{figure}

The index properties, smoothness and squintability, numerically
characterise the complexity of projection pursuit index, and hence
potential optimisation problems.
Table~\ref{tbl-smoothness-squintability} presents the parameters for
calculating both metrics estimated from the Gaussian process (variance,
range, smooth, and nugget) and the scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column \(\nu\) is used as the smoothness metric and the
column \(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

\hypertarget{tbl-smoothness-squintability}{}
\begin{table}
\caption{\label{tbl-smoothness-squintability}Parameters estimated from the Gaussian process (outputscale \(\eta\),
lengthscale \(\ell\), smoothness \(\nu\), and nugget \(\sigma\)) and
scaled logistic function (\(\theta_1\) to \(\theta_4\) and
\(\varsigma\)) for the pipe-finding and sine-wave finding problems. The
columns \(\nu\) and \(\varsigma\) represent the smoothness and
squintability measures respectively. }\tabularnewline

@@

\begin{tabular}{|>{}lcc>{}c|ccc>{}c|cccc>{}c|}
\toprule
  & shape & index & d & $\eta$ & $\ell$ & $\nu$ & $\sigma$ & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & $\varsigma$\\
\midrule
1 & pipe & holes & 6 & 0.04 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.01 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.01 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.01 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.13 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.13 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.35 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.35 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor & 6 & 0.18 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess & 6 & 0.28 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines & 6 & 0.20 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

\hypertarget{tbl-mod-output}{}
\begingroup
\fontsize{12.0pt}{14.4pt}\selectfont
\setlength{\LTpost}{0mm}
\begin{longtable}{lcc}
\caption{\label{tbl-mod-output}Jellyfish success rate relative to index properties and jellyfish
hyper-parameters. This is the summary from a logistic regression fit to
smoothness, squintability, dimension, running time, number of jellyfish
and maximum number of tries. Interestingly, squintability and dimension
strongly affect jellyfish optimisation success. The number of jellies
marginally affects success, but index smoothness, running longer and
increasing the number of tries do not. }\tabularnewline

\toprule
\textbf{Characteristic} & \textbf{OR} \textbf{(95\% CI)}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Smoothness & 3.30 (0.08 to 274) & 0.533 \\ 
Squintability & 7.85 (2.40 to 37.7) & {\bfseries 0.003} \\ 
Dimension & 0.53 (0.30 to 0.83) & {\bfseries 0.014} \\ 
Long runtime & 0.42 (0.03 to 5.44) & 0.497 \\ 
Number of jellyfish & 1.24 (0.98 to 1.63) & {\bfseries 0.089} \\ 
Maximum number of tries & 1.12 (0.84 to 1.52) & 0.452 \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}OR = Odds Ratio, CI = Confidence Interval\\
\end{minipage}
\endgroup

\hypertarget{sec-discussion}{%
\section{Practicalities}\label{sec-discussion}}

The simulation study demonstrated that these metrics can be used to
assess a PP index, and compare optimisers. Here we explain how they can
be used for new index and optimiser development. Using the JSO optimiser
for PPGT requires a change in user behaviour. For all the existing
optimisers a single optimisation path is followed. The JSO has multiple
optimisation paths, and needs additional tools to incorporate into the
PPGT, as explained here.

\hypertarget{using-the-jso-in-a-ppgt}{%
\subsection{Using the JSO in a PPGT}\label{using-the-jso-in-a-ppgt}}

@@

\hypertarget{supplementary-materials}{%
\section*{Supplementary materials}\label{supplementary-materials}}
\addcontentsline{toc}{section}{Supplementary materials}
